\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-05-19 doc/publications/plos16/nm_alt.tex

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1,2},
Piotr Szul\textsuperscript{3},
Robert Dunne\textsuperscript{4}, 
Paul Leo \textsuperscript{5},
Emma L. Duncan \textsuperscript{5,6,7},
Natalie A. Twine\textsuperscript{1,8}
Stephanie Li\textsuperscript{9},
James Doecke\textsuperscript{9},
Nick Ellis\textsuperscript{10}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} ANU, Canberra, Australia
\\
\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
\\
\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
\\
\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
\\
\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
\\
\textbf{8} MQ, QLD, Australia
\\
\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\bigskip

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the biomedical space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests \cite{Breiman2001} are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
trees on different compute nodes and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages to date are implemented in R \cite[]{R-language}, although it has been
reported that current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
This is despite R now supporting long vectors (length greater than $2^{31}-1$), and the \randomforest implementation in
R lending itself to parallelisation by growing multiple trees simultaneously and combining them for the final
result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide'' genomic data have been developed in
other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go, achieves fast run times by
effective use of the CPU cache, optimising for different classes of features and efficiently multithreading.  Similarly,
\ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is specifically optimised for dealing with
large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed.  
The programs are limited to utilising only CPUs that are on the same computer node (multithreading) or
farm out independent tasks to CPUs distributed across nodes that do not require communication between the
processes (a separate tree is grown on each node).
%else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
%processes (a separate tree is grown on each node).  
Hadoop/Spark overcomes these limitations by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies. In the case of
random forests, the computations for each node of a tree can hence be handed off to separate processors (cores).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview of Random Forests}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the data sets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Sets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site \cite[]{1KG2012}.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Results
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with bone mineral density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

\section{Conclusion}

Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
We have demonstrated that for data sets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a bone mineral density study. 
We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.

@@

>>>>

% This is a separate version of the CursedForest manuscript following Nature Method formatting guidelines. A Word document has been created based on this. See the Word document for the latest version of the paper.
%
% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Machine Learning Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1,2},
Piotr Szul\textsuperscript{3},
Robert Dunne\textsuperscript{4}, 
Paul Leo \textsuperscript{5},
Emma L. Duncan \textsuperscript{5,6,7},
Natalie A. Twine\textsuperscript{1,8}
Stephanie Li\textsuperscript{9},
James Doecke\textsuperscript{9},
Nick Ellis\textsuperscript{10}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} ANU, Canberra, Australia
\\
\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
\\
\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
\\
\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
\\
\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
\\
\textbf{8} MQ, QLD, Australia
\\
\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\bigskip

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of ``big" and ``wide'' data is especially
pronounced in the biomedical space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests \cite{Breiman2001} are well
suited for processing ``wide'' genomic data for two reasons: firstly, while other machine learning applications have
the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting; secondly, random
forests are easy to parallelise, as the forest is a sum of decision trees, it is possible to grow separate
trees on different compute nodes and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed.  
The programs are limited to utilising only CPUs that are on the same computer node (multithreading) or
farm out independent tasks to CPUs distributed across nodes that do not require communication between the
processes (a separate tree is grown on each node).
%else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
%processes (a separate tree is grown on each node).  
Hadoop/Spark overcomes these limitations by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
random forests, the computations for each node of a tree can hence be handed off to separate processors.

@@

\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with bone mineral density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

\section{Discussion}
Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
We have demonstrated that for data sets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a bone mineral density study. 
We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview of Random Forests (This should go to Supplemental Information)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest (This should go to Results section)}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the data sets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Sets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.
